---
title: "NIVA_Gadus"
author: "Hanne Gladhaug Woien Mo"
date: "2023-11-20"
output: html_document
editor_options: 
  chunk_output_type: console
---
## Examining data from NIVA
Data for Atlantic cod *Gadus morhua* was extracted from the database of previously collected values from other NIVA-projects. 
```{r}
library(readxl)
library(dplyr)
library(ggplot2)
library(tidyverse)
Gadus <- read_excel("~/NIVA - EXPECT/Data/Gadus.xlsx", 
    sheet = "Export Worksheet")
View(Gadus)
```
The data consists of information on carbon content, measured consenrtations of pollutants and isotopes of nitrogen etc. The fish has been collected from different locations along the Norwegian coast - listed in the column *Description*.

```{r}
#Grouping by the parameter column to get an overview of the data.
parameters <- unique(Gadus$REPORTED_NAME) %>%
  sort()

parameters
```

The parameter that is interesting for the model is lipid content and weigths of organs. *Fettinnhold* is the interesting parameter. 

```{r}
#Filtering the data to include only the lipid content. 

fat <- Gadus %>% 
  filter(REPORTED_NAME == "Fettinnhold") %>% 
  group_by("TISSUE")

#Sorting into different tissues
muscle <- fat%>% 
  filter(TISSUE == "MU-Muskel") 
liver <- fat %>% 
  filter(TISSUE == "LI-Lever")

hist(muscle$NUMERIC_ENTRY)

hist_liver <- ggplot(liver, aes(x = NUMERIC_ENTRY)) +
  geom_histogram(binwidth = 1, fill = "lightblue", color = "blue", alpha = 0.7) +
  labs(title = "Fat content in liver",
       x = "Fat content (%)",
       y = "Frequency") +
  theme_minimal()

print(hist_liver)

hist_muscle <- ggplot(muscle, aes(x = NUMERIC_ENTRY)) +
  geom_histogram(binwidth = 1, fill = "lightblue", color = "blue", alpha = 0.7) +
  labs(title = "Fat content in liver",
       x = "Fat content (%)",
       y = "Frequency") +
  theme_minimal()

print(hist_muscle)
```

```{r}
liver_var <- var(liver$NUMERIC_ENTRY, na.rm = TRUE)
#This variance is very high, some of the metadata 
boxplot_liver <- ggplot(liver, aes(x = AQUAMONITOR_NAME, y = NUMERIC_ENTRY)) +
  geom_boxplot(fill = "violet") +
  labs(title = "Fat content in liver by age", 
       x = "Location",
       y = "Fat content in liver") + 
  coord_flip()
print(boxplot_liver) 
```

```{r}
variance_by_location <- liver %>%
  group_by(AQUAMONITOR_NAME) %>%
  summarize(variance = var(NUMERIC_ENTRY), sd = sd(NUMERIC_ENTRY))

view(variance_by_location)

variance_by_project <- liver %>% 
  group_by(PROSJEKT) %>% 
  summarize(variance = var(NUMERIC_ENTRY), sd = sd(NUMERIC_ENTRY))

view(variance_by_project)
```


